### 2020-02-17 chapters/background.tex
Aggregations,  Aggregations properties, Sliding Window Aggregations (SWAGs), SWAG algorithms

@@

Whether working with a static bounded data set or a dynamic data stream bounded by windows, there might be a need to reduce the collection of data to a single value. The operation of reducing a collection of items to a single value is known by many names: fold, reduce, accumulate or aggregate. Aggregate operations are present in most programming languages and frameworks like Apache Hadoop and its MapReduce paradigm. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm relies on applying a map and reduce functions to the collection of data, in this order. The map task will transform each item and the reduce task will merge the collection into a single output. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregation state. Doing this recursively for each item in the collection results in a single value: the aggregation value for the data set.

Aggregations over data are fundamental in overall data processing, not just stream processing. There are multiple examples of aggregations, such as computing counts, averages, maxima or minima and more. In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus aggregating data in the sliding windows. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).

Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand and later on examine this aggregation, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

@@

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and a corresponding value a list of temperatures, from the past 5 minutes. Finally, we compute an \textit{AVG} aggregation for each of the racks, resulting in the average temperature per rack in the past 5 minutes.

%The distinction between the aggregation algorithm and the aggregation operation is important. For example, the operation we are interested in might be a sum, an average, a maximum, a minimum or something more complex such as a membership test or the count of distinct elements. The aggregation algorithm comes with its own data structures for storing the intermediate aggregation state and a specific algorithm to update this data structure. For example, consider the computation of an aggregation like a sum over data grouped by a true sliding window of any size. One way of computing said sum would be to, after each unitary step, compute from scratch the sum of all window values. However, a more efficient way of doing so would be to compute the total sum of the window once and then, at each step, evict the oldest element and subtract it from the total sum state and include the new event in the window while adding it up to the aggregation state. In this scenario, even though the aggregation operator is the same --- \textit{i.e.} a sum --- the algorithms differ in that one recalculates everything from scratch and the other incrementally changes the aggregation state. Hence, the operation and the algorithm differ greatly from each other.

@@

In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions. Both these functions are first introduced in the work of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. The \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

@@

    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter
\end{enumerate}

These properties are important to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations operations. This way, the multiple SWAG algorithms presented in Chapter 3 will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation operator properties.

\subsection{Algorithms for Sliding Window Aggregations}

@@

>>>>

An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations have an exact value or state in that there is no error or loss of resolution associated. Approximate aggregations have an error margin to the associated aggregation value. This error is usually controllable by parameter optimization.

Aggregate operations are present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying a map function to the collection of data and then a reduce one. The map task will transform each item as desired and the reduce task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregation state. Doing this recursively for each item in the collection results in a single value: the aggregation value for the data set.

@@

In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus being able to aggregate the data in the sliding window. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand and later on examine this aggregation, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

@@

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

@@

In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions. The \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

@@

    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

These properties are used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this Thesis will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.

\subsection{Sliding Window Aggregation Algorithms}

@@

A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window. A SWAG algorithm may restrict the set of possible aggregations based on their properties. Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state. Despite storing information using different structures, every SWAG algorithm must implement three methods:

\begin{itemize}
    \item \textit{insert} - inserts new value in the window
    \item \textit{evict} - evicts all values that no longer fit in the window
    \item \textit{query} - returns the aggregation state/value
\end{itemize}

Tangwongsan et al. present two of the most basic SWAG algorithms: \textit{Recalculate From Scratch (RFS)} and \textit{Subtract On Evict (SOE)}.

\subsubsection{Recalculate From Scratch}
Recalculate From Scratch (RFS) is the simplest and most general SWAG algorithm. As the name suggests, this algorithm recomputes the aggregation over the entire window for each insertion and eviction. For instance, the computation of a sum over a sliding window using RFS means that upon each insertion or eviction all elements are summed.

This approach does not restrain the aggregation in terms of properties: it works for every family of aggregations. However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.

\subsubsection{Subtract On Evict}
Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} function is invertible. That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.

SOE inserts new elements into the window and combines the previous aggregation value with the new element. On eviction, SOE applies the inverse of the combine function, $\ominus$, to expire the evicted element from the aggregation state.

With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.

\subsection{Summary}
This subsection adds nothing new to the previous ones. However, it concisely summarizes the most relevant previously discussed concepts.

An aggregation is an operation that reduces a collection of values to a single one. Examples of such are sums, minima, averages or finding the number of distinct elements in a set (cardinality). An aggregation may be exact or approximate. Exact aggregations have an exact value or state in that there is no error or loss of resolution associated. Approximate aggregations have an error margin to the associated aggregation value. This error is usually controllable by parameter optimization.

Sliding Window Aggregation algorithms allow the computation of exact or approximate aggregations over a data stream in a sliding window fashion. Each algorithm must implement three methods: \textit{insert}, \textit{evict} and \textit{query}. A SWAG algorithm may restrict the set of possible aggregations based on their properties.
